---
title: "Data Preparation"
author: "Suicide Squad"
date: 14 Feb 2023
date-modified: "`r Sys.Date()`"
execute: 
  echo: true
  eval: true
  warning: false
format: html
editor: visual
---

## Step-by-Step Data Preparation

### 1. Installing and launching required R packages

```{r}
pacman::p_load("rworldmap", "tidyverse")
```

### 2. Loading the data

```{r}
suicidedata <- list.files(path = "data",
                          pattern = "IHME-GBD_2019_DATA",
                          full.names = TRUE) %>%
  lapply(read_csv) %>%
  bind_rows

world_region <- read_csv("data/world-regions-according-to-the-world-bank.csv")
```

```{r}
str(suicidedata)
```

### 3. Data wrangling

-   Filter for "Self=harm"

-   Select relevant columns and arrange rows based on location_name

-   Pivot_wider by metric_name (number, percent, and rate)

-   Calculate the population size, death number and death rate

```{r}
suicidedata_cln <- suicidedata %>%
  filter(cause_name == "Self-harm") %>%
  arrange(location_name) %>%
  select(4,6,8,10,(12:14)) %>%
  pivot_wider(names_from = metric_name, values_from = val, names_glue = "suicide_{metric_name}") %>%
  mutate(population_size = suicide_Number/(suicide_Rate/100000),
         death_number = suicide_Number/suicide_Percent,
         death_rate = death_number/population_size*100000)
```

### 4. Matching country's name with region (ISO3 data)

Identify key differences

```{r}
suicide_vs_region <- setdiff(suicidedata_cln$location_name, world_region$Entity) %>%
  enframe(name = NULL, value = "desn") %>% 
  arrange(desn)
```

```{r}
unique(suicidedata_cln$location_name)
```

```{r}
#the list of countries in suicidedata_clean but not in world_map
suicide_vs_region  %>% 
  knitr::kable(caption = "Countries in suicide dataset vs. ISO3 regions: Coverage diff",
               row.names = TRUE)
```

#### 4.1 Changing the names of countries

Change the names of countries to match the two datasets

```{r}
suicidedata_cln_ctry <- suicidedata_cln %>%
  rename(country = location_name) %>%
  mutate(country = case_when(country == "Bolivia (Plurinational State of)" ~ "Bolivia",
                             country == "Brunei Darussalam" ~ "Brunei",
                             country == "Cabo Verde" ~ "Cape Verde",
                             country == "Côte d'Ivoire" ~ "Cote d'Ivoire",
                             country == "Democratic People's Republic of Korea" ~ "North Korea",
                             country == "Democratic Republic of the Congo" ~ "Democratic Republic of Congo",
                             country == "Iran (Islamic Republic of)" ~ "Iran",
                             country == "Lao People's Democratic Republic" ~ "Laos",
                             country == "Micronesia (Federated States of)" ~ "Micronesia",
                             country == "Republic of Korea" ~ "South Korea",
                             country == "Republic of Moldova" ~ "Moldova",
                             country == "Russian Federation" ~ "Russia",
                             country == "Syrian Arab Republic" ~ "Syria",
                             country == "Taiwan (Province of China)" ~ "Taiwan",
                             country == "United Republic of Tanzania" ~ "Tanzania",
                             country == "Venezuela (Bolivarian Republic of)" ~ "Venezuela",
                             country == "Viet Nam" ~ "Vietnam",
                             TRUE ~ country))

world_region_cln <- world_region %>%
  rename(country = Entity) %>%
  mutate(country = case_when(country == "Timor" ~ "Timor-Leste",
                             country == "United States" ~ "United States of America",
                             TRUE~country))

### Progress check
setdiff(suicidedata_cln_ctry$country, world_region_cln$country)  %>% 
  enframe(name = NULL, value = "diff") %>% 
  knitr::kable(caption = "Remaining Cases", 
               row.names = TRUE)
```

Joining the two datasets

```{r}
suicidedata_cln_region = suicidedata_cln_ctry %>%
  left_join(world_region_cln, by = c('country' = 'country')) %>%
  select(!Year) %>%
  rename(code = Code,
         region = World_RegionbyWorld_Bank) %>%
  relocate(code:region, .after = country)

suicidedata_cln_region
```

### 5. Matching country's name with world_map

rworldmap - Example code below

https://cran.r-project.org/web/packages/rworldmap/vignettes/rworldmap.pdf

```{r}
suicidedata_cln_map <- joinCountryData2Map(suicidedata_cln_region,
                                 joinCode = "ISO3",
                                 nameJoinColumn = "code")
```

Other References using typical world map - https://rpubs.com/Thom_JH/798825
