---
title: "EDA: Analysis of Suicide Rate with Other Parameters"
author: "Magdalene"
execute: 
  warning: false
---

# Load Packages and Data Set

```{r}
pacman::p_load(tidyverse, ExPanDaR, knitr, ggstatsplot)

suicide_analysis <- read.csv('data/suicidedata_analysis.csv', header = T)
```

# Visualising Missing Values

```{r}
missing.values <- suicide_analysis %>%
    pivot_longer(cols = SR_AS_M:mental_disorder, 
                 names_to = "rate",
                 values_to = "values") %>%
    mutate(is.missing = is.na(values)) %>%
    group_by(rate, is.missing) %>%
    summarise(num.missing = n()) %>%
    filter(is.missing==T) %>%
    select(-is.missing) %>%
    arrange(desc(num.missing)) 

missing.values %>% kable()
```

```{r}
prepare_missing_values_graph(df = suicide_analysis
                               |> select(region, year,
                                         gdp_per_capita:mental_disorder), 
                               ts_id = "region")
```

```{r}
prepare_missing_values_graph(df = suicide_analysis 
                             |> select(year, gdp_per_capita:mental_disorder), 
                               ts_id = "year")
```

As there is a large amount of missing values between the years 1990 to 2005 for a few variables, we will filter the data down to only include values from 2005 onwards. We will also exclude the variable "Average hours worked" since there is a consistently high proportion of missing values.

```{r}
suicide_analysis_filtered <- suicide_analysis %>%
  filter(year %in% seq(2005,2020,1)) %>%
  select(!avg_hrs_worked) %>%
  select(region, country, year, SR_AS_T, gdp_per_capita:mental_disorder)

prepare_missing_values_graph(df = suicide_analysis_filtered, 
                             ts_id = "region")
```

This seems to bring down the proportion of missing values, but there remains areas and variables with a high number of missing values. Let's take a look at how the missing values are distributed within countries in each region.

```{r}
missing_val_graph <- function(df, filter_col, filter_val){
  prepare_missing_values_graph(df = df 
                             |> filter({{filter_col}} == filter_val),
                               ts_id = "country")
}
```

```{r}
missing_val_graph(suicide_analysis_filtered, region, "Sub-Saharan Africa")
```

```{r}
missing_val_graph(suicide_analysis_filtered, region, "Middle East and North Africa")
```

```{r}
missing_val_graph(suicide_analysis_filtered, region, "South Asia")
```

```{r}
missing_val_graph(suicide_analysis_filtered, region, "East Asia and Pacific")
```

```{r}
missing_val_graph(suicide_analysis_filtered, region, "Europe and Central Asia")
```

```{r}
missing_val_graph(suicide_analysis_filtered, region, "North America")
```

```{r}
missing_val_graph(suicide_analysis_filtered, region, "Latin America and Caribbean")
```

# Missing Values Graph v2 after exclusions

```{r}
exclude_countries <- c('South Sudan', 'Somalia', 'Seychelles', 'Eritrea', 
                       'Equitorial Guinea', 'Yemen', 'Syria', 'Libya',
                       'Vanuatu', 'Tuvalu', 'Tokelau', 'Taiwan', 'Palau',
                       'Northern Mariana Islands', 'North Korea', 'Niue',
                       'Nauru', 'Marshall Islands', 'Guam', 'Cook Islands',
                       'American Samoa', 'San Marino', 'Monaco', 'Greenland',
                       'Andorra', 'Bermuda', 'United States Virgin Islands', 
                       'Saint Kitts and Nevis', 'Puerto Rico', 'Grenada', 
                       'Dominica', 'Cuba', 'Bahamas', 'Antigua and Barbuda',
                       'Zimbabwe')

missing_val_graph_v2 <- function(df, filter_col, filter_val){
  prepare_missing_values_graph(df = df 
                               |> filter(!(country %in% exclude_countries))
                               |> filter({{filter_col}} == filter_val),
                               ts_id = "country")
}
```


```{r}
missing_val_graph_v2(suicide_analysis_filtered, region, "Sub-Saharan Africa")
```

```{r}
missing_val_graph_v2(suicide_analysis_filtered, region, "Middle East and North Africa")
```

```{r}
missing_val_graph_v2(suicide_analysis_filtered, region, "South Asia")
```

```{r}
missing_val_graph_v2(suicide_analysis_filtered, region, "East Asia and Pacific")
```

```{r}
missing_val_graph_v2(suicide_analysis_filtered, region, "Europe and Central Asia")
```

```{r}
missing_val_graph_v2(suicide_analysis_filtered, region, "North America")
```

```{r}
missing_val_graph_v2(suicide_analysis_filtered, region, "Latin America and Caribbean")
```

# Comparison between Factors

```{r}
suicide_analysis_factors <- suicide_analysis_filtered %>%
  select(gdp_per_capita:mental_disorder)

ggcorrmat(data = suicide_analysis_factors,
          ggcorrplot.args = list(lab_size = 2)) + 
  theme(axis.text.x = element_text(size = 7, angle = 90, vjust = 0.5, hjust=1),
        axis.text.y = element_text(size =7, vjust = 0.3)) 
```

```{r}
correl_plot <- function(df, col_x, col_y){
  ggscatterstats(data = df
                 |> filter(!is.na({{col_x}})) 
                 |> filter(!is.na({{col_y}})), 
                 x = {{col_x}}, y = {{col_y}}, 
                 type = "nonparametric") + 
      theme(axis.text.x = element_text(size = 7, angle = 90, vjust = 1, hjust=1),
            axis.text.y = element_text(size =7)) +
      labs(title = paste0("Correlation between ", deparse(substitute(col_x)), 
                      " and ", deparse(substitute(col_y))))
}
```

```{r}
correl_plot(suicide_analysis_filtered, gdp_per_capita, SR_AS_T)
```

```{r}
correl_plot(suicide_analysis_filtered, hdi, SR_AS_T)
```

```{r}
correl_plot(suicide_analysis_filtered, depr_rate, SR_AS_T)
```

```{r}
correl_plot(suicide_analysis_filtered, gini, SR_AS_T)
```

```{r}
correl_plot(suicide_analysis_filtered, violent_death_rate, SR_AS_T)
```

```{r}
correl_plot(suicide_analysis_filtered, pop.extreme_poverty, SR_AS_T)
```

```{r}
correl_plot(suicide_analysis_filtered, unemployment_rate, SR_AS_T)
```

```{r}
correl_plot(suicide_analysis_filtered, HAQ, SR_AS_T)
```

```{r}
correl_plot(suicide_analysis_filtered, healthcare_spend, SR_AS_T)
```

```{r}
correl_plot(suicide_analysis_filtered, DALYs_NCD, SR_AS_T)
```

```{r}
correl_plot(suicide_analysis_filtered, happiness, SR_AS_T)
```
